Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-02-15 draft.tex

The SN Ia rate as a function of stellar mass and the best fitting model for SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much stronger correlation with mass and make up nearly all SNe in high-mass hosts.

>>>>

The SN Ia rate as a function of stellar mass SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR. The corresponding DTD fit results in a steep power-law for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope for low-stretch SNe representing a much more delayed population. Such a distinction has been predicted by \citet{Rigault2013,Childress2014,Rigault2018,Nicolas2020} based on the intrinsic $x_1$ distribution and its observed relationship with stellar mass and age.

From the differing DTDs that describe SNe with low and high stretch values we infer that there are either multiple channels through which SN Ia explode which occur on differing timescales, or that the explosion mechanism evolves with progenitor age. Two scenarios, that are not necessarily mutually exclusive, that could lead to different DTD decay rates are as follows. 
The first such scenario is to assume that all SNe Ia come from the same initial population of stars, and thus that the progenitors of both low and high stretch SNe form at the same time. It is then necessary to invoke models in which the WD progenitors of low-stretch SNe evolve on longer timescales than those of high-stretch SNe, for example due to a different initial binary separation distribution or accretion rate.

Alternatively, low- and high-stretch SNe could form via identical evolutionary channels but with a different onset time since the episode of star-formation. The predominantly high-stretch "prompt" SNe would begin exploding as soon as the WDs are formed (nominally 40 Myr) whereas the mainly low-stretch "tardy" SNe would only begin exploding after an extended period of time of the order 1 Gyr. The rates of both populations would then both follow the fiducial $t^{-1}$ distribution beginning at this onset time.
Such a scenario has been explored theoretically, primarily in the sub-Chandrasekhar mass (sub-$M_{\mathrm{Ch}}$) paradigm \citep[e.g.][]{Sim2010,Blondin2017,Shen2017}, usually associated with various forms of the DD scenario. Sub-$M_{\mathrm{Ch}}$ SNe Ia typically involve the merger of two WDs where the SN luminosity is correlated with the mass of the primary WD; a correlation between primary WD mass and age would then lead to the different rates, and apparent DTDs, of low and high-stretch SNe observed in this work.

We explore this scenario by fitting the stretch-separated sub-samples with modified DTDs. We fix the DTD to a power-law with index $-1$, and fix the normalisation to the best fitting value from the full sample. We model the DTD as comprising two components: the prompt, high-stretch SNe and the tardy, low stretch SNe, and we introduce two timescales, $t_1$ and $t_2$. At $\tau\leq t_1$, the DTD is caused solely by prompt SNe; at $\tau \geq t_2$ only tardy SNe explode. In between where $t_1 \leq\tau\leq t_2$, the DTD is a sum of DTD (prompt) and DTD (tardy), where we model the relative fraction $f$ as a smooth linear slope between $t_1$ and $t_2$:
\begin{equation}
    f_{\mathrm{tardy}} = \frac{\tau - t_1}{t_2 - t_1}\,,
\end{equation}
and
\begin{equation}
    f_{\mathrm{prompt}} = 1 - f_{\mathrm{tardy}}\,.
\end{equation}
With $\beta$ and $A$ fixed,  we fit the split-$x_1$ SN Ia rate with $t_1$ and $t_2$ as free parameters, and weak normal priors $p(t_1) \sim \mathcal{N}\left(0.5~\mathrm{Gyr}, 0.5~\mathrm{Gyr}\right)$ $p(t_2) \sim \mathcal{N}\left(1~\mathrm{Gyr}, 0.5~\mathrm{Gyr}\right)$ and the constraint $t_2 > t_1$.

\subsubsection{The late end of the DTD \label{subsubsec:subtypes}}

Despite the reasonable fit of the evolving population model, the fit to tardy SNe diverges from the data at high stellar mass, corresponding to the oldest average stellar age. It is difficult to reconcile this turnover with the simple DTD models used thus far in this work, as it would require a steepening or complete turn-off of the DTD at late ($\tau\gtrsim 5$~Gyr) times. Here we describe two possible ways in which the model can be reconciled with the data in the highest-mass galaxies.

One possibility is that sub-luminous SNe are more numerous in such predominantly old stellar populations, such that the fraction passing our light curve cuts is lower than the rest of the sample. Such a phenomenon could be caused by SN Ia sub-classes such as SN 1991bg-like SNe, which are known to explode exclusively in old environments \citep{Perets2010} at large delay times $>6$~Gyr \citep{Panther2019}.
